Nucleotide Diversity Loss on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-11-02 manuscript.tex
Update manuscript.tex

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked diversity during plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that average neutral polymorphism levels for genes on the Y chromosome are \textasciitilde 2.1\% the value of their homologues on the X chromosome, indicating a significant departure from neutrality and a chromosome-wide reduction of \textasciitilde 93\%. We demonstrate that the magnitude of this diversity loss is inconsistent with neutral models that account for the occurrence of female-biased sex ratios, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations and estimates of the distribution of fitness effects of deleterious mutations, we show that neutral polymorphism on the Y can be explained by the effects of linked purifying selection, although selective sweeps may have also contributed. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study suggests that the early stages of Y chromosome degeneration are characterized by strong selective interference among many linked sites rather than gene silencing followed by neutral genetic drift.
%abstract is already 270 words. Should be 250 (but might be ok). Just can't get longer.
\end{abstract}

>>>>

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male reproductive success can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked diversity during plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that average neutral polymorphism levels for genes on the Y chromosome are \textasciitilde 2.1\% the value of their homologues on the X chromosome, indicating a significant departure from neutrality and a chromosome-wide reduction of \textasciitilde 93\%. We demonstrate that the magnitude of this diversity loss is inconsistent with neutral models that account for the occurrence of female-biased sex ratios, or by reduced male $N_{e}$ caused by high variance in male reproductive success. Rather, using forward simulations and estimates of the distribution of fitness effects of deleterious mutations, we show that neutral polymorphism on the Y can be explained by the effects of linked purifying selection, although selective sweeps may have also contributed. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study suggests that the early stages of Y chromosome degeneration are characterized by strong selective interference among many linked sites rather than primarily gene silencing followed by neutral genetic drift.
%abstract is already 270 words. Should be 250 (but might be ok). Just can't get longer.
\end{abstract}
